Using the internal language of toposes in commutative algebra

### 2017-09-15 slides-pssl101.tex
Slides for PSSL 101 (WIP)

\title{First steps in synthetic algebraic geometry}
\author{Ingo Blechschmidt}
\date{July 21st, 2017}

@@

  \hil{First steps in \\ synthetic algebraic geometry}
  \medskip

@@

  Oberseminar Algebra/Geometrie/Topologie \\
  Heinrich-Heine-Universität Düsseldorf \\
  \ \\
  July 21st, 2017
  \par
\end{frame}

\backupstart
{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/unterbacher-see}}
\begin{frame}[plain]
  \vspace{\textheight}\vspace{-0em}
  \centering
  \textcolor{white}{\emph{Camping at Unterbacher See}} \\
  \par
\end{frame}}

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
\frame[t]{\frametitle{Quick summary}
  By employing the internal language of toposes in various ways, you can
  pretend that:
  \begin{enumerate}
    \item Reduced rings are Noetherian and in fact fields.
    \item Sheaves of modules are plain modules.
    \item Schemes are sets: \[ \PP^2_S = \{
    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
    0 \}. \]
  \end{enumerate}
}
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

@@

@@

\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
  \begin{block}{Crucial property: Locality}
    If~$U = \bigcup_i U_i$, then
      $U \models \varphi$ iff $U_i \models
      \varphi$ for all~$i$.
  \end{block}

  \begin{block}{Crucial property: Soundness}
    If~$U \models \varphi$ and if $\varphi$
    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no axiom of choice
    }},
    then~$U \models \psi$.
  \end{block}

@@

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{2.9cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{2.9cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

@@

\section[Little Zariski]{The little Zariski topos of a scheme}

\frame[t]{\frametitle{The little Zariski topos}
  \begin{block}{Definition}The \hil{little Zariski topos} of a scheme~$X$ is the
  category~$\Sh(X)$ of set-valued sheaves on~$X$.\end{block}

>>>>

\title{Using the internal language of toposes in commutative algebra}
\author{Ingo Blechschmidt}
\date{September 17st, 2017}

@@

  \hil{Using the internal language of toposes \\ in commutative algebra}
  \medskip

@@

  PSSL 101 in Leeds \\
  \ \\
  September 17st, 2017
  \par
\end{frame}

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
\frame[t]{\frametitle{Quick summary}
  By employing the internal language of toposes in various ways, you can
  pretend that:
  \begin{enumerate}
    \item Sheaves of modules are plain modules.
    \item Schemes are sets: \[ \PP^2_S = \{
    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
    0 \}. \]
    \item Reduced rings are Noetherian and in fact fields.
  \end{enumerate}
}
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

@@

\frame[t]{\frametitle{The little Zariski topos}
  Let~$A$ be a ring.
  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.
  \hil{Yes,} if we allow a change of topos:
  Then $A \to \O_{\Spec(A)}$ is the universal localisation.

  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}

  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
  for some~$g_i$.
}

% Wieso hat LocRing --> Ring keinen Linksadjungierten?
% Sollte also nicht Limiten bewahren.

% Recall the definition of a local ring and a local homomorphism!

@@

\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
  \begin{block}{Locality}
    If~$U = \bigcup_i U_i$, then
      $U \models \varphi$ iff $U_i \models
      \varphi$ for all~$i$.
  \end{block}

  \begin{block}{Soundness}
    If~$U \models \varphi$ and if $\varphi$
    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no axiom of choice
    }},
    then~$U \models \psi$.
  \end{block}

@@

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{2.9cm} iff $f|_U = 0 \in \F(U)$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{2.9cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

@@

\section[Little Zariski]{The little Zariski topos of a ring}

\begin{frame}{XXX}
\end{frame}

% O_{Spec(A)} as the reification of all stalks
% O_{Spec(A)} integral domain iff all stalks integral domains
% Transfer theorem

% *Constructed* to yield a universal localisation (mention classifying topos),
% *observed* to fulfil further properties.
% Give Mulvey/Tierney citation, state background (quasicoherence)
% also neg neg (f = 0) ==> f = 0

% Anonymous Noetherian condition
% Prove it

% Grothendieck's generic freeness lemma

% Outlook

\frame[t]{\frametitle{The little Zariski topos}
  Let~$A$ be a commutative ring (in~$\Set$).

  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.

  \hil{Yes,} if we allow a change of topos:
  Then $A \to \O_{\Spec(A)}$ is the universal localisation.

  \begin{block}{Definition}The \hil{little Zariski topos} of~$A$ is the
  category~$\Sh(\Spec(A))$ of set-valued sheaves on the \hil{spectrum} of~$A$.\end{block}

  The topology on~$\Spec(A)$ is generated by open subsets~$D(f)$ for~$f \in A$
  such that~$\Spec(A) = \bigcup_i D(f_i)$ if and only if~$1 = \sum_i g_i f_i$
  for some~$g_i$.
